---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    fig_width: 5
    keep_md: yes
    number_sections: yes
    toc: yes
---

## Loading and preprocessing the data
If required file is not present in our working directory, then download zip file, extract it to WD and delete temp file.

```{r}
if(!file.exists("activity.csv")) {
        # create temporary file with random name
        temp <- tempfile()
        # download zip archive
        download.file("http://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip", temp)
        # extract it
        unzip(temp)
        # delete temporary file
        unlink(temp)
        # clear memory (delete temp variable)
        rm(temp)
}

# read data
data <- read.csv("activity.csv")
```

## Explore the dataset
Get some sense on our data

```{r}
str(data)
summary(data)
head(data)
tail(data)
```

## What is mean total number of steps taken per day?



## What is the average daily activity pattern?



## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?
