---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    fig_width: 5
    keep_md: yes
    number_sections: yes
    toc: yes
---

## Load the data
If required file is not present in our working directory:
* download zip file
* extract it to WD
* delete temp file
* clear memory
* read data

```{r}
if(!file.exists("activity.csv")) {
        # create temporary file with random name
        temp <- tempfile()
        # download zip archive
        download.file("http://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip", temp)
        # extract it
        unzip(temp)
        # delete temporary file
        unlink(temp)
        # clear memory (delete temp variable)
        rm(temp)
}

# read data
df <- read.csv("activity.csv")
```

## Explore the data
* number, names, types of variables
* number of observations including missing values
* basic stats: mins, maxes, etc
* see some actual data

```{r}
str(df)
summary(df)
head(df)
tail(df)
```

## Preprocess the data
* convert 'date' variable to correct class (Factor -> Date)

* remove observations with NA values

```{r}
df$date <- as.Date(df$date, "%Y-%m-%d")
df.tidy <- na.omit(df)
str(df.tidy)
```

## What is mean total number of steps taken per day?
* calculate total number of steps per day
* show some new data
* create histogram

```{r}
steps <- aggregate(steps ~ date, df, sum)
head(steps)
hist(steps$steps,
     breaks = 10,
     xlab = "Number of steps",
     xlim = c(0, 25000),
     ylim = c(0, 25), 
     main = "Total number of steps a day")
```

* calculate **mean** and **median** of the number of steps
```{r}
mean(steps$steps)
median(steps$steps)
```

Mean  : `r mean(steps$steps)`
Median: `r median(steps$steps)`

## What is the average daily activity pattern?

* make a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)
```{r}
# average number of steps for each interval for all days
interval <- aggregate(steps ~ interval, df.tidy, mean)

# plot the average number of steps per day by interval
plot(interval$interval, interval$steps,
     lwd  = 2,
     xaxt = "n",
     yaxt = "n",
     type = "l",
     col  = "blue",
     xlab = "Interval",
     ylab = "Average Number of Steps",
     main = "Average Number of Steps per Day by Interval")
axis(1, at = seq(100, 2500, by = 100), las = 2)
axis(2, las = 2)
```

* find 5-minute interval that contains the maximum of steps, on average across all days
```{r}
max <- interval[which.max(interval$steps),]
```

Interval number `r max$interval` contains maximum number of steps - `r as.integer(max$steps)`

## Imputing missing values

Note that there are a number of days/intervals where there are missing values (coded as NA).
The presence of missing days may introduce bias into some calculations or summaries of the data.

1. Calculate and report the total number of missing values in the dataset
```{r}
sum(is.na(df))
```

The number of NA's is **`r sum(is.na(df))`**.

2. Devise a strategy for filling in all of the missing values in the dataset
NA values will be replaced by the mean number of steps per identical 5-minute interval

3. Create a new dataset with the missing data filled in
```{r}
# replace NA values by the mean number of steps per identical 5-minute interval
for (i in 1:nrow(df)) {
    if(is.na(df$steps[i]))
        df$steps[i] <- interval$steps[interval$interval == df$interval[i]]
}
```

4. Make a histogram of the total number of steps taken each day and calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment?
What is the impact of imputing missing data on the estimates of the total daily number of steps?
```{r}
# calculate updated total number of steps 
steps.new <- aggregate(steps ~ date, df, sum)

# build histogram
hist(steps.new$steps,
     col = "blue",
     breaks = 10,
     xlab = "Number of steps",
     xlim = c(0, 25000),
     ylim = c(0, 25), 
     main = "Total number of steps a day")

# calculate new mean and median values
mean(steps.new$steps)
median(steps.new$steps)
```



## Are there differences in activity patterns between weekdays and weekends?
